Investigation of ant cuticle dataset using image texture analysis

### 2022-08-29 paper/main.tex
address comments

\def\typeofarticle{Research Article} 
\def\currentvolume{3} 
\def\currentissue{1}
\def\currentyear{2022} 
\def\currentmonth{} 
\def\ppages{xx--xx} 
\def\DOI{xxxx} 
\def\Received{xx xx 2022} 
\def\Revised{xx xx 2022}
\def\Accepted{xx xx 2022} 
\def\Published{xx 2022}

@@

\begin{abstract}
    Ant cuticle texture presumably provides some type of function, and therefore
    is useful to research for ecological applications and bioinspired designs.
    In this study, we employ statistical image texture analysis and deep machine
    learning methods to classify similar ant species based on morphological
    features. We establish a public database of ant cuticle images for research.
    We provide a comparative study of the performance of image texture
    classification and deep machine learning methods on an ant cuticle dataset.
    Our results show that the deep learning methods give higher accuracy than
    statistical methods in recognizing ant cuticle textures. Our experiments
    also reveal that the deep learning networks designed for image texture
    performs better than the general deep learning networks.
\end{abstract}
\keywords{texture analysis; image processing; classification; machine learning;
    ant cuticle images; ecology}
\maketitle

@@

Classical image texture analysis methods can be grouped into four categories:
statistical, structural, model-based, and transform-based methods
\cite{bharati_image_2004}. The statistical method is often used by feature
extraction methods and then fed into the classification algorithm. Markov random
field models have also been used for textural image interpretation
\cite{hassner_use_1981, cross_markov_1983}. Transform-based methods use some
functions to decompose an image texture into a set of basic feature images.
Gabor filters and Wavelet expansions are two of the widely used approaches
\cite{bovik_multichannel_1990}.

@@

In the past decade, deep learning using CNNs has emerged as the mainstream
technology for image analysis. Following this trend, various related network
architectures have been designed to specifically characterize textural images.
Among these, the Fisher-vector CNN descriptor (FV-CNN) proposed by Cimpoi et al.
\cite{cimpoi_deep_2015} is widely accepted as one of the most important
pioneering works. It applies FV pooling to deep features obtained via a CNN
pre-trained using the ImageNet \cite{krizhevsky_imagenet_2017} to obtain encoded
features for texture classification. FV-CNN is capable of achieving much
improved classification accuracy comparing to traditional hand-crafted texture
features. However, CNN-based deep learning is only used in its feature
extraction module. Its encoding and classification modules still use traditional
learning methods, thus not supporting an end-to-end deep learning. In the
context of texture image classification, an end-to-end deep learning model
refers to a neural network architecture which consists of all the modules
(feature extraction, encoding and classification) between the initial input data
and the final output classification result, with all the modules being trained
simultaneously.

@@

Compared with the traditional method in which a kernel must be designed by an
engineer for extracting features, the deep learning networks can automatically
extract the features through the training. In addition, the deep learning
networks achieve the higher accuracy than that of the classical approaches,
although the deep networks require much more data for training.

@@

Feng et al. \cite{feng_automated_2013} apply an automated system to classify
moth images based on semantic related visual attributes, which are defined as a
pattern on the moth wings. Feng et al. \cite{feng_automated_2013} use a custom
texture descriptor based on the combination of GLCM and \textit{scale-invariant
    feature transform} (SIFT) features \cite{gotlieb_texture_1990,
    lowe_distinctive_2004}. The method is used to classify 50 different moth species
across 8 families \cite{feng_automated_2013}. Their results
\cite{feng_automated_2013} suggest that traditional feature extraction
techniques for the semantic visual attributes of the moth wings are sufficient
for training a classifier to classify an image between 10 randomly selected moth
species.

@@

The completed dataset\footnote{All code is publicly available on GitHub
    \url{https://github.com/ngngardner/cuticulus}} contains 2,499 images as a
4-dimensional array with shape of (samples, rows, columns, channels), of
which there are 3 channels extracted from the RGB images. 1072 samples of
rough textured ant cuticle textures comprise 43\% of the dataset. The
remaining 1427 samples of smooth textured ant cuticle textures comprise 57\%
of the dataset. Due to the variety of the ant head image attributes, we
apply simple preprocessing before the images are used in our model. We want
the images to have a uniform size for simplicity in our classification
process. Since the ant head images are typically centered in the image, we
apply a center crop to each image to create a square image of the same size.
Once the image is square, we resize each image to a fixed size of 256x256
pixels. We leave other discrepancies in the images untouched. A summary of
the resulting dataset is shared in Table \ref{tab:dataset-summary}.

\begin{table}[h]
    \centering
    \caption{Summary of the dataset.}
    \begin{tabular}{|l|c|}
        \toprule
        Total images  & 2499    \\
        Rough images  & 1072    \\
        Smooth images & 1427    \\
        Image size    & 256x256 \\
        \bottomrule
    \end{tabular}
    \label{tab:dataset-summary}
\end{table}

\section{Methods for comparative analysis and experimental setup}

@@

The K-means algorithm is one of widely used clustering algorithms in the pattern
recognition \cite{lloyd_least_1982}. It is a single pixel based classification
algorithm for images. The K-views is an algorithm that uses several
characteristic views for the classification of images \cite{hung_use_2002}. The
K-views algorithm is suitable for classifying image textures that have basic
local patterns repeated in a periodic manner. In contrast to the K-means,
K-views looks at the neighboring pixels for providing the spatial relationship
for classification.  Several variations of the basic K-views model have been
proposed \cite{yang_image_2003, lan_improved_2010}.

For the feasibility study in this ant image datasets, we use both K-means and
K-views in our experiments as both are statistical clustering methods. The
experiments may indicate how well the statistical approach will do in this type
of textural images. In pre-processing the dataset, we apply the Gabor filter to
transform the gray-scale ant image into four Gabor-filtered channels, subtract
the gray-scale image from each channel, take the absolute value of the resulting
images, and then normalize them. This pre-processing will highlight some of
characteristics features such as edge-like in the image. This step is very
useful in feeding the datasets to the K-views algorithms. A 4-channel Gabor
filter is applied in the pre-processing step for the K-means algorithm. The
theta $\theta$ parameters in the Gabor filters used are multiples of $\pi/4$
radians starting from $0$ to $\pi$ radians. The Gabor filter lambda $\lambda$
parameter is equal to $\pi$.

@@

Our first deep learning model is \textit{visual geometry group} (VGG), a
convolutional neural network that takes advantage of very small convolutional
filters in a deep network architecture \cite{simonyan_very_2015}. We compare
four architectures of VGG: VGG11, VGG13, VGG16, and VGG19. The primary
difference between the architectures is the number of layers in each model. Our
second model is \textit{residual network} (ResNet), a deep network architecture
that includes shortcut connections between layers (residual connections)
\cite{he_deep_2015}. We compare three architectures of ResNet: ResNet18,
ResNet50, and ResNet101. Again, the primary difference between the architectures
is the number of layers in each model.

For our ResNet models, we have two versions: randomized and pretrained. The
randomized version is the same architecture, but the weights are randomly
initialized. The pretrained version has weights from training on the CIFAR
dataset, an image dataset with 1000 classes. In this case, we are fine-tuning
the pretrained model. For VGG, we are only using the randomized version. The
base VGG architecture also has an output layer of size 1000. Since we are
working with a binary classification problem, we modify the architecture for all
models to have an output layer of size 2. Each model is trained over 100 epochs,
using stochastic gradient descent with momentum. The batch size is set to 16
images. We apply a learning rate of 0.001 and momentum parameter of 0.9.

@@

In addition to the classical texture analysis methods and the deep learning
models for general image analysis, we also examine deep learning analysis
methods specifically designed for textural images. For this examination, we
adopt the deep residual pooling (DRP) network developed by Mao et al.
\cite{mao_deep_2021}. The DRP framework consists of a unique residual pooling
layer, which is formed by a residual encoding module followed by an aggregation
module. The residual encoding module extracts relevant spatial information,
while the aggregation module performs averaging to obtain orderless
low-dimension features for classification.

According to \cite{mao_deep_2021}, the DRP network can be used either with a
single residual pooling layer, applied right before the classifier layer, or
with multiple residual pooling layers, applied to the output of multiple
convolutional layers before feeding the combined pooling results into the
classifier layer. Additionally, an auxiliary classifier layer may also be used.
In our experiments, we follow these strategies. For each scenario, we conduct
two experiments. One experiment employs weights that are randomly initialized,
and the other uses weights that are fine tuned over those from a pretrained
model.

To handle the imbalance of the dataset, we apply undersampling for each class for
the training dataset. By using random stratified sampling, we construct a
training set with 800 images per class. The remaining images are randomly split
between test and validation, which turns out to roughly a 60\%/20\%/20\% train,
test, and validation data split. With 272 rough samples and 627 smooth samples
left over after the stratified split, the test dataset has roughly 136 rough
samples and 313 smooth samples. Since these leftover samples are split with code
by 50\% there will be some rounding variance and therefore the test dataset
built at run-time will not always have exactly the same number of samples.

@@

We evaluate the performance of the models according to standard evaluation
methods. Since we are working with a binary classification problem, we use a
standard confusion matrix to evaluate the accuracy, precision, and F1 score. We
also apply Grad-CAM with manual inspection to visualize the activation weights
for classified images to visualize which features lead to the classification
result. Finally, we apply t-SNE to visualize the separation learned for the
model to further analyze the classifications made by the model.

@@

Experiments are run on an Ubuntu 18.04 LTS Lambda Labs GPU server. The server
contains 8 NVIDIA GeForce RTX 2080 Ti graphics cards with 12GB of memory each.
The server uses an Intel Xeon Silver 4116 with 48 total threads and maximum
frequency of 3.000 GHz, and has 256GB of RAM.

\section{Results and analysis}

@@

To begin, we share the results on the each algorithm specified in the
methodology section using the evaluation metrics described. The results are
shown in Table \ref{tab:results}. It should be noted that due to the class
imbalance in the dataset, the F1 score is the preferable metric to the accuracy.
% Therefore, the results have been sorted by F1 score in descending order.

@@

\FloatBarrier
The results in the previous sections show that the fine-tuned ResNet models
outperform the VGG and randomly initialized ResNet models on the task of ant
head image classification. Additionally, the DRP models perform better than the
the general deep learning models VGG and ResNet on the task of ant head image
classification.The fine-tuned DRP single-layer model with auxiliary classifier
performed the best with an average F1 score of 0.92. We further analyze the
separation learned by both ResNet101 models in the following section.

@@

\begin{figure}[h]
    \centering
    \begin{subfigure}{.45\textwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/plots/resnet101_gt_tsne.png}
        % \caption{Ground truth labels}
    \end{subfigure}
    \begin{subfigure}{.45\textwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/plots/resnet101_pred_tsne.png}
        % \caption{Predicted labels}
    \end{subfigure}
    \caption{t-SNE visualization of the embeddings of the second to last layer
        of the randomly initialized ResNet101 model trained on ant head image
        dataset.}
    \label{fig:resnet101_tsne}
\end{figure}

In this section, we provide visualization of the fine-tuned ResNet101 model and
the randomly initialized ResNet101 model using t-SNE dimensionality reduction.
First, we run the dataset preprocessing method and initialize both models. Then,
both models are trained according to the training parameters and the state of
each model is saved. To visualize the deep extracted features, we modify each
model to obtain the embeddings of the second to last layer. Then, we use the
t-SNE algorithm to reduce the dimensionality of the embeddings to 2 dimensions.
We plot side-by-side the ground truth and predicted labels for each model.
Figure \ref{fig:resnet101_tsne} shows the results of the trained randomly
initialized model and Figure \ref{fig:fresnet101_tsne} shows the results of the
fine-tuned model. Based on the visual results of the t-SNE visualization, we can see that the
fine-tuned model learned a stronger separation of the two classes, which
reinforces the results that the fine-tuned model received a higher average
accuracy.

\begin{figure}[h]
    \centering
    \begin{subfigure}{.45\textwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/plots/fresnet101_gt_tsne.png}
        % \caption{Ground truth labels}
    \end{subfigure}
    \begin{subfigure}{.45\textwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/plots/fresnet101_pred_tsne.png}
        % \caption{Predicted labels}
    \end{subfigure}
    \caption{t-SNE visualization of the embeddings of the second to last layer
        of the fine-tuned ResNet101 model trained on ant head image dataset.}
    \label{fig:fresnet101_tsne}
\end{figure}

@@

Next, we provide some visual analysis of some correctly and
incorrectly classified images using GradCAM. We provide two categories and two
subcategories in our analysis. The two categories are correct and incorrect
classification. Regardless of the feature activation map, the correctly
classified images have the same predicted label as the ground truth, and
incorrectly classified images have different predicted labels. The two
subcategories are expected and non-expected feature activation. In the expected
case, the features that are used to compute the classification are the same as
the features used by the assistants in the sculpture identification process. In
general, the features used by the assistants are the textures of the cuticle on
the ant head. In the non-expected case, the features used to compute the
classification are not from the head, for example, from the background,
extraneous text, or the body of the ant. We used randomly selected images from
the dataset and the fine-tuned ResNet101 model to perform the analysis. We show
the GradCAM results in Figures \ref{fig:correct_images} and
\ref{fig:incorrect_images}. The left image shows the preprocessed image input
to the model. The right image shows the GradCAM output based on the
classification. Four specimens were selected randomly from each category and
subcategory.

@@

\newcommand{\subwidth}{0.35\textwidth}
% \begin{figure}
%     \centering
%     \begin{subfigure}{\subwidth}
%         \includegraphics[width=1\linewidth]{thesis_assets/gradcam/correct_ideal/518.png}
%         \caption*{Smooth texture}
%         \label{fig:correct_ideal_518}
%     \end{subfigure}
%     \begin{subfigure}{\subwidth}
%         \includegraphics[width=1\linewidth]{thesis_assets/gradcam/correct_ideal/808.png}
%         \caption*{Smooth texture}
%         \label{fig:correct_ideal_808}
%     \end{subfigure}
%     \begin{subfigure}{\subwidth}
%         \includegraphics[width=1\linewidth]{thesis_assets/gradcam/correct_ideal/842.png}
%         \caption*{Rough texture}
%         \label{fig:correct_ideal_842}
%     \end{subfigure}
%     \begin{subfigure}{\subwidth}
%         \includegraphics[width=1\linewidth]{thesis_assets/gradcam/correct_ideal/1091.png}
%         \caption*{Rough texture}
%         \label{fig:correct_ideal_1091}
%     \end{subfigure}
%     \caption{Correctly classified images using expected features.}
%     \label{fig:correct_ideal}
% \end{figure}

% \begin{figure}
%     \centering
%     \begin{subfigure}{\subwidth}
%         \includegraphics[width=1\linewidth]{thesis_assets/gradcam/correct_nonideal/346.png}
%         \caption*{Rough texture}
%         \label{fig:correct_nonideal_346}
%     \end{subfigure}
%     \begin{subfigure}{\subwidth}
%         \includegraphics[width=1\linewidth]{thesis_assets/gradcam/correct_nonideal/1554.png}
%         \caption*{Rough texture}
%         \label{fig:correct_nonideal_1554}
%     \end{subfigure}
%     \begin{subfigure}{\subwidth}
%         \includegraphics[width=1\linewidth]{thesis_assets/gradcam/correct_nonideal/1694.png}
%         \caption*{Rough texture}
%         \label{fig:correct_nonideal_1694}
%     \end{subfigure}
%     \begin{subfigure}{\subwidth}
%         \includegraphics[width=1\linewidth]{thesis_assets/gradcam/correct_nonideal/388.png}
%         \caption*{Smooth texture}
%         \label{fig:correct_nonideal_388}
%     \end{subfigure}
%     \caption{Correctly classified images using non-expected features.}
%     \label{fig:correct_nonideal}
% \end{figure}

\begin{figure}
    \centering
    \begin{subfigure}{\subwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/gradcam/correct_ideal/518.png}
        \caption*{Smooth texture}
        \label{fig:correct_ideal_518}
    \end{subfigure}
    \begin{subfigure}{\subwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/gradcam/correct_ideal/808.png}
        \caption*{Smooth texture}
        \label{fig:correct_ideal_808}
    \end{subfigure}
    \begin{subfigure}{\subwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/gradcam/correct_ideal/842.png}
        \caption*{Rough texture}
        \label{fig:correct_ideal_842}
    \end{subfigure}
    \begin{subfigure}{\subwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/gradcam/correct_ideal/1091.png}
        \caption*{Rough texture}
        \label{fig:correct_ideal_1091}
    \end{subfigure}
    \begin{subfigure}{\subwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/gradcam/correct_nonideal/346.png}
        \caption*{Rough texture}
        \label{fig:correct_nonideal_346}
    \end{subfigure}
    \begin{subfigure}{\subwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/gradcam/correct_nonideal/1554.png}
        \caption*{Rough texture}
        \label{fig:correct_nonideal_1554}
    \end{subfigure}
    \begin{subfigure}{\subwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/gradcam/correct_nonideal/1694.png}
        \caption*{Rough texture}
        \label{fig:correct_nonideal_1694}
    \end{subfigure}
    \begin{subfigure}{\subwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/gradcam/correct_nonideal/388.png}
        \caption*{Smooth texture}
        \label{fig:correct_nonideal_388}
    \end{subfigure}
    \caption{Correctly classified images from fine-tuned ResNet101.}
    \label{fig:correct_images}
\end{figure}

>>>>

\def\typeofarticle{Research Article} 
\def\currentvolume{2} 
\def\currentissue{2}
\def\currentyear{2022} 
\def\currentmonth{} 
\def\ppages{xx--xx} 
\def\DOI{xxxx} 
\def\Received{xx xx 2022} 
\def\Revised{xx xx 2022}
\def\Accepted{xx xx 2022} 
\def\Published{xx 2022}

@@

\begin{abstract}
    Ant cuticle texture presumably provides some type of function, and therefore
    is useful to research for ecological applications and bioinspired designs.
    In this study, we employ statistical image texture analysis and deep machine
    learning methods to classify similar ant species based on morphological
    features. We establish a public database of ant cuticle images for research.
    We provide a comparative study of the performance of image texture
    classification and deep machine learning methods on this ant cuticle dataset.
    Our results show that the deep learning methods give higher accuracy than
    statistical methods in recognizing ant cuticle textures. Our experiments
    also reveal that the deep learning networks designed for image texture
    performs better than the general deep learning networks.
\end{abstract}
\keywords{texture analysis; image processing; classification; machine learning;
    ant cuticle images; ecology}
\maketitle

@@

Classical image texture analysis methods can be grouped into four categories:
statistical, structural, model-based, and transform-based methods
\cite{bharati_image_2004}. The statistical method is often used for image
classification after feature extraction. Markov random field models have also
been used for textural image interpretation \cite{hassner_use_1981,
    cross_markov_1983}. Transform-based methods use some functions to decompose an
image texture into a set of basic feature images. Gabor filters and wavelet
expansions are two of the widely used approaches \cite{bovik_multichannel_1990}.

@@

In the past decade, deep learning using CNNs has emerged as the mainstream
technology for image analysis. Following this trend, various related network
architectures have been designed to specifically characterize textural images.
Among these, the Fisher-vector CNN descriptor (FV-CNN) proposed by Cimpoi et al.
\cite{cimpoi_deep_2015} is widely accepted as one of the pioneering works. It
applies Fisher-vector pooling to deep features obtained via a CNN pre-trained
using the ImageNet \cite{krizhevsky_imagenet_2017} to obtain encoded features
for texture classification. FV-CNN is capable of achieving higher classification
accuracy comparing to traditional hand-crafted texture features. However,
CNN-based deep learning is only used in its feature extraction module. Its
encoding and classification modules still use traditional learning methods, thus
not supporting an \textit{end-to-end} deep learning. In the context of texture
image classification, an end-to-end deep learning model refers to a neural
network architecture which consists of all the modules (feature extraction,
encoding and classification) between the initial input data and the final output
classification result, with all the modules being trained simultaneously.

@@

Compared with the traditional method in which a kernel must be designed by an
engineer for extracting features, the deep learning networks can automatically
extract the features through the training. In addition, the deep learning
networks can 6achieve the higher accuracy than that of the classical approaches,
although the deep networks require much more data for training.

@@

Feng et al. \cite{feng_automated_2013} apply an automated system to classify
moth images based on semantic related visual attributes, which are defined as a
pattern on the moth wings. Feng et al. \cite{feng_automated_2013} use a custom
texture descriptor based on the combination of GLCM and \textit{scale-invariant
    feature transform} (SIFT) features \cite{gotlieb_texture_1990,
    lowe_distinctive_2004}. The method is used to classify 50 different moth species
across 8 families \cite{feng_automated_2013}. Their results
\cite{feng_automated_2013} suggest that traditional feature extraction
techniques for the semantic visual attributes of the moth wings are sufficient
for training a classifier to classify an image between 10 randomly selected moth
species. Marques et al. \cite{marques_ant_2018} propose an ensemble-based method
to classify a large number of ant genera. They use ant images sourced from AntWeb
in order to classify 44,806 specimens with at least 1 picture into 57 genera.

@@

The completed dataset\footnote{The dataset is available on GitHub
    \url{https://github.com/ngngardner/cuticulus}} contains 2,499 images as a
4-dimensional array with shape of (samples, rows, columns, channels), of
which there are 3 channels extracted from the RGB images. 1072 samples of
rough textured ant cuticle textures comprise 43\% of the dataset. The
remaining 1427 samples of smooth textured ant cuticle textures comprise 57\%
of the dataset. Due to the variety of the ant head image attributes, we
apply simple preprocessing before the images are used in our model. We want
the images to have a uniform size for simplicity in our classification
process. Since the ant head images are typically centered in the image, we
apply a center crop to each image to create a square image of the same size.
Once the image is square, we resize each image to a fixed size of 256 $\times$ 256
pixels. We leave other discrepancies in the images untouched. A summary of
the resulting dataset is shared in Table \ref{tab:dataset-summary}.

\begin{table}[h]
    \centering
    \caption{Summary of the dataset.}
    \begin{tabular}{|l|c|}
        \toprule
        Total images  & 2499             \\
        Rough images  & 1072             \\
        Smooth images & 1427             \\
        Image size    & 256 $\times$ 256 \\
        \bottomrule
    \end{tabular}
    \label{tab:dataset-summary}
\end{table}

\section{K-views, K-means, and deep learning models for experiments}

@@

The K-means algorithm is one of widely used clustering algorithms in the pattern
recognition \cite{lloyd_least_1982}. It is a single pixel based classification
algorithm for images. Based on the concept of K-means, K-views was developed,
which uses characteristic views for image texture representation
\cite{hung_use_2002}. The K-views algorithm is suitable for classifying image
textures that have basic local patterns repeated in a periodic manner. In
contrast to the K-means, K-views looks at the neighboring pixels for providing
the spatial relationship for classification.  Several variations of the basic
K-views model have been proposed \cite{yang_image_2003, lan_improved_2010}.

For the feasibility study in this ant image datasets, we tested the statistical
clusterings methods and deep learning methods separately. Both K-means and
K-views were used for statistical clustering methods. The experiments may
indicate how well the statistical approach and deep learning networks will do in
this type of textural images. In pre-processing the dataset for the statistical
clustering methods, we apply the Gabor filter to transform the gray-scale ant
image into four Gabor-filtered channels, subtract the gray-scale image from each
channel, and take the absolute value of the resulting images. This
pre-processing will highlight some of characteristic features in textures, such
as edge-like in the image. This step is very useful in feeding the datasets to
the K-views algorithms. The theta $\theta$ parameters in the Gabor filters used
are multiples of $\pi/4$ radians starting from $0$ to $\pi$ radians. The Gabor
filter lambda $\lambda$ parameter is equal to $\pi$.

@@

The first deep learning model used in our experiment is \textit{visual geometry
    group} (VGG), a convolutional neural network that takes advantage of very
small convolutional filters in a deep network architecture
\cite{simonyan_very_2015}. We compare four architectures of VGG11, VGG13,
VGG16, and VGG19. The primary difference between the architectures is the
number of layers in each model. The second deep learning model used in our
experiments is \textit{residual network} (ResNet), a deep network
architecture that includes shortcut connections between layers (residual
connections) \cite{he_deep_2015}. We compare three architectures of
ResNet18, ResNet50, and ResNet101. Again, the primary difference between the
architectures is the number of layers in each model.

For the initial ResNet models, we have two versions: randomized and pretrained.
The randomized version is the same architecture, but the weights are randomly
initialized. The pretrained version has weights from training on the CIFAR
dataset, an image dataset with 1000 classes. Since the pretrained weights are
readily available, we also evaluate a fine-tuned model. For VGG, we only use the
randomized version. The base VGG architecture also has an output layer of size
1000. Since we are working with a binary classification problem, we modify the
architecture for all models to have an output layer of size 2. Each model is
trained over 100 epochs, using stochastic gradient descent with momentum. The
batch size is set to 16 images. We apply a learning rate of 0.001 and momentum
parameter of 0.9.

@@

In addition to the classical texture analysis methods and the deep learning
models for general image analysis, we also examine deep learning analysis
methods specifically designed for textural images. For this examination, we
adopt the deep residual pooling (DRP) network \cite{mao_deep_2021}. The DRP
framework consists of a unique residual pooling layer, which is formed by a
residual encoding module followed by an aggregation module. The residual
encoding module extracts relevant spatial information, while the aggregation
module performs averaging to obtain orderless low-dimension features for
classification.

According to \cite{mao_deep_2021}, the DRP network can be used either with a
single residual pooling layer, applied right before the classifier layer, or
with multiple residual pooling layers, applied to the output of multiple
convolutional layers before feeding the combined pooling results into the
classifier layer. Additionally, an auxiliary classifier layer may also be used.
In our experiments, we follow the same strategy. For each scenario, we conduct
two experiments. One experiment employs weights that are randomly initialized,
and the other uses weights that are fine tuned over those from a pretrained
model.

To handle the imbalance of the dataset, we apply undersampling for each class
for the training dataset \cite{tiittanen_novel_2022}. By using random stratified
sampling, we construct a training set with 800 images per class. The remaining
images are randomly split between test and validation, which turns out to
roughly a 60\%/20\%/20\% train, test, and validation data split. With 272 rough
samples and 627 smooth samples left over after the stratified split, the test
dataset has roughly 136 rough samples and 313 smooth samples. Since these
leftover samples are split with code by 50\% there will be some rounding
variance and therefore the test dataset built at run-time will not always have
exactly the same number of samples.

@@

We evaluate the performance of the models using accuracy, precision, and F1
score. We also apply Grad-CAM with manual inspection to visualize the activation
weights for classified images to show which features lead to the
classification result. Finally, we apply t-SNE to visualize the separation
learned for the model to further analyze the classifications made by the model.

\section{Experimental results and analysis}

@@

Experiments were run on an Ubuntu 18.04 LTS Lambda Labs GPU server. The server
contained 8 NVIDIA GeForce RTX 2080 Ti graphics cards with 12GB of memory each.
The server used an Intel Xeon Silver 4116 with 48 total threads and maximum
frequency of 3.000 GHz, and has 256GB of RAM. K-means algorithm comes from the
scikit-learn\footnote{\url{https://scikit-learn.org/}} library. The K-views
implementation was based on the K-means implementation from scikit-learn. VGG
and ResNet models come from the
torchvision\footnote{\url{https://pytorch.org/vision}} library. DRP model code
was tweaked from the original author's uploaded code on
GitHub\footnote{\url{https://github.com/maoshangbo/DRP-Texture-Recognition}}.

@@

We share the results on each algorithm specified in Section 4. Both the K-means
and K-views algorithms are unsupervised. All of the deep learning methods used
are supervised. The results are shown in Table \ref{tab:results}. It should be
noted that due to the class imbalance in the dataset, the F1 score is the
preferable metric to the accuracy.
% Therefore, the results have been sorted by F1 score in descending order.

@@

\FloatBarrier

The results show that the fine-tuned ResNet models outperform the VGG and
randomly initialized ResNet models on the task of ant head image classification.
The DRP models perform better than the the general deep learning models VGG and
ResNet on the task of ant head image classification. The fine-tuned DRP
single-layer model with auxiliary classifier performed the best with an average
F1 score of 0.92. We further analyze the separation learned by both ResNet101
models in the following section.

@@

\begin{figure}[h]
    \centering
    \begin{subfigure}{.45\textwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/plots/resnet101_gt_tsne.png}
        % \caption{Ground truth labels}
    \end{subfigure}
    \begin{subfigure}{.45\textwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/plots/resnet101_pred_tsne.png}
        % \caption{Predicted labels}
    \end{subfigure}
    \caption{t-SNE visualization of the embeddings of the second to last layer
        of the randomly initialized ResNet101 model trained on dataset.}
    \label{fig:resnet101_tsne}
\end{figure}

In this section, we provide visualization of the fine-tuned ResNet101 model and
the randomly initialized ResNet101 model using t-SNE dimensionality reduction.
To visualize the deep extracted features, we modify each model to obtain the
embeddings of the second to last layer. We plot side-by-side the ground truth
and predicted labels for each model. Figure \ref{fig:resnet101_tsne} shows the
results of the trained randomly initialized model and Figure
\ref{fig:fresnet101_tsne} shows the results of the fine-tuned model. Based on
the visual results of the t-SNE visualization, we can see that the fine-tuned
model learned a stronger separation of the two classes, which reinforces the
results that the fine-tuned model received a higher average accuracy.

\begin{figure}[h]
    \centering
    \begin{subfigure}{.45\textwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/plots/fresnet101_gt_tsne.png}
        % \caption{Ground truth labels}
    \end{subfigure}
    \begin{subfigure}{.45\textwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/plots/fresnet101_pred_tsne.png}
        % \caption{Predicted labels}
    \end{subfigure}
    \caption{t-SNE visualization of the embeddings of the second to last layer
        of the fine-tuned ResNet101 model trained on dataset.}
    \label{fig:fresnet101_tsne}
\end{figure}

@@

Next, we provide some visual analysis of some correctly and incorrectly
classified images using GradCAM. In the expected case, the features that are
used to compute the classification are the same as the features used by the
assistants in the sculpture identification process. In general, the features
used by the assistants are the textures of the cuticle on the ant head. In the
non-expected case, the features used to compute the classification are not from
the head, for example, from the background, extraneous text, or the body of the
ant. We used randomly selected images from the dataset and the fine-tuned
ResNet101 model to perform the analysis. We show the GradCAM results in Figures
\ref{fig:correct_images} and \ref{fig:incorrect_images}. The left image shows
the preprocessed image input to the model. The right image shows the GradCAM
output based on the classification. Four specimens were selected randomly from
each category and subcategory.

@@

\newcommand{\subwidth}{0.35\textwidth}
\begin{figure}
    \centering
    \begin{subfigure}{\subwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/gradcam/correct_ideal/518.png}
        \caption*{Smooth texture}
        \label{fig:correct_ideal_518}
    \end{subfigure}
    \begin{subfigure}{\subwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/gradcam/correct_ideal/808.png}
        \caption*{Smooth texture}
        \label{fig:correct_ideal_808}
    \end{subfigure}
    \begin{subfigure}{\subwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/gradcam/correct_ideal/842.png}
        \caption*{Rough texture}
        \label{fig:correct_ideal_842}
    \end{subfigure}
    \begin{subfigure}{\subwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/gradcam/correct_ideal/1091.png}
        \caption*{Rough texture}
        \label{fig:correct_ideal_1091}
    \end{subfigure}
    \begin{subfigure}{\subwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/gradcam/correct_nonideal/346.png}
        \caption*{Rough texture}
        \label{fig:correct_nonideal_346}
    \end{subfigure}
    \begin{subfigure}{\subwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/gradcam/correct_nonideal/1554.png}
        \caption*{Rough texture}
        \label{fig:correct_nonideal_1554}
    \end{subfigure}
    \begin{subfigure}{\subwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/gradcam/correct_nonideal/1694.png}
        \caption*{Rough texture}
        \label{fig:correct_nonideal_1694}
    \end{subfigure}
    \begin{subfigure}{\subwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/gradcam/correct_nonideal/388.png}
        \caption*{Smooth texture}
        \label{fig:correct_nonideal_388}
    \end{subfigure}
    \caption{GradCAM visualization for correctly classified images from
        fine-tuned ResNet101. The red sections of the heatmap show which features
        had higher activation for the classication.}
    \label{fig:correct_images}
\end{figure}

\begin{figure}
    \centering
    \begin{subfigure}{\subwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/gradcam/incorrect_ideal/84.png}
        \caption*{Rough texture}
        \label{fig:incorrect_ideal_84}
    \end{subfigure}
    \begin{subfigure}{\subwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/gradcam/incorrect_ideal/86.png}
        \caption*{Smooth texture}
        \label{fig:incorrect_ideal_86}
    \end{subfigure}
    \begin{subfigure}{\subwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/gradcam/incorrect_ideal/177.png}
        \caption*{Smooth texture}
        \label{fig:incorrect_ideal_177}
    \end{subfigure}
    \begin{subfigure}{\subwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/gradcam/incorrect_ideal/197.png}
        \caption*{Smooth texture}
        \label{fig:incorrect_ideal_197}
    \end{subfigure}
    \begin{subfigure}{\subwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/gradcam/incorrect_nonideal/1.png}
        \caption*{Rough texture}
        \label{fig:incorrect_nonideal_1}
    \end{subfigure}
    \begin{subfigure}{\subwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/gradcam/incorrect_nonideal/22.png}
        \caption*{Rough texture}
        \label{fig:incorrect_nonideal_22}
    \end{subfigure}
    \begin{subfigure}{\subwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/gradcam/incorrect_nonideal/61.png}
        \caption*{Rough texture}
        \label{fig:incorrect_nonideal_61}
    \end{subfigure}
    \begin{subfigure}{\subwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/gradcam/incorrect_nonideal/204.png}
        \caption*{Rough texture}
        \label{fig:incorrect_nonideal_204}
    \end{subfigure}
    \caption{GradCAM visualization for incorrectly classified images from
        fine-tuned ResNet101. The red sections of the heatmap show which features
        had higher activation for the classication.}
    \label{fig:incorrect_images}
\end{figure}
